---
title: "Power Analysis for Two-Sample t-Tests: A Worked Example"
author: "Jane Smith"
date: "2026-05-08"
categories: [methods, R, statistics]
description: "A reproducible walkthrough of sample size and power calculations for two-sample t-tests, with simulated power curves across common effect sizes."
bibliography: references.bib
---

## Introduction

Determining adequate sample size before data collection is a cornerstone of
rigorous experimental design. The required sample size depends on the
expected effect size, the desired power $(1 - \beta)$, and the significance
level $\alpha$. This post provides a reproducible walkthrough using R's
built-in `power.t.test` function.

We follow the framework of [@cohen1988], which introduced the widely-used
conventions for small ($d = 0.2$), medium ($d = 0.5$), and large ($d = 0.8$)
effect sizes.

## Sample size formula

For a balanced two-sample t-test, the minimum sample size per group is
determined by:

$$
n = \frac{2\sigma^2(z_{1-\alpha/2} + z_{1-\beta})^2}{\delta^2}
$$

where $\delta$ is the true mean difference, $\sigma^2$ is the common
variance, $z_{1-\alpha/2}$ is the critical value for a two-tailed test at
level $\alpha$, and $z_{1-\beta}$ is the critical value for power
$1 - \beta$.

## Setup

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(dplyr)
```

## Power curves by effect size

The plot below shows how statistical power changes as a function of sample
size for three canonical effect sizes.

```{r}
#| label: fig-power
#| fig-cap: "Power curves for a two-sample t-test (α = 0.05, two-tailed) across small (d = 0.2), medium (d = 0.5), and large (d = 0.8) Cohen's d effect sizes."

effect_sizes <- c(0.2, 0.5, 0.8)
n_seq <- seq(5, 200, by = 5)

power_grid <- expand.grid(n = n_seq, d = effect_sizes) |>
  rowwise() |>
  mutate(
    power = power.t.test(
      n = n, delta = d, sd = 1,
      sig.level = 0.05, type = "two.sample",
      alternative = "two.sided"
    )$power
  ) |>
  ungroup() |>
  mutate(effect = factor(d, labels = c("Small (d = 0.2)",
                                        "Medium (d = 0.5)",
                                        "Large (d = 0.8)")))

ggplot(power_grid, aes(x = n, y = power, color = effect)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "grey40") +
  annotate("text", x = 195, y = 0.82, label = "80% power",
           hjust = 1, size = 3.2, color = "grey40") +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 1)) +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Sample size per group (n)",
       y = "Power",
       color = "Effect size") +
  theme_minimal(base_size = 13)
```

## Minimum n for 80% power

```{r}
#| label: tbl-summary
#| tbl-cap: "Minimum sample size per group required to achieve 80% power at α = 0.05 (two-tailed two-sample t-test)."

min_n_table <- tibble(
  `Effect size (d)` = c(0.2, 0.5, 0.8),
  Label = c("Small", "Medium", "Large")
) |>
  rowwise() |>
  mutate(
    `Min n per group` = ceiling(
      power.t.test(
        delta = `Effect size (d)`, sd = 1,
        sig.level = 0.05, power = 0.80,
        type = "two.sample", alternative = "two.sided"
      )$n
    )
  ) |>
  ungroup()

knitr::kable(min_n_table)
```

Detecting a small effect ($d = 0.2$) requires substantially more
participants than detecting a large effect — a practical reminder that
underpowered studies are most likely to miss precisely the effects that are
hardest to measure.

## References
